---
title: "Simulation4"
author: "Leona"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview
This R script is designed to simulate a dataset with various socioeconomic variables, apply different methods to generate missing data, and create new variables based on specified functions and relationships. The script is structured to systematically create data, introduce missingness in a controlled way, and explore the relationships between variables. The results are summarized using statistical models and tables to evaluate the impacts of these simulated processes.

# Initial set-up
This section: 
- loads necessary libraries, 
- sets a random seed for reproducibility, 
- and defines the number of data points (n = 10000) to be simulated.
```{r}
# load libraries 
library(tidyverse)
library(MASS)
library(dplyr)

# set seed 
set.seed(294)

# set number of data points
n <- 10000
```

# Set-up for data traits
This code block sets up parameters for adding noise to specific variables (noise_income, noise_house_ownership, and noise_crime), which introduces variability into the dataset. It also defines whether a non-linear relationship should be introduced between income and sensitive attributes or education (non_linear_income). Additionally, it includes a flag (missing_data) to indicate whether missing values should be introduced into the dataset.

The code also includes a parameter to control the non-linearity of the decision boundary (non_linear). Lastly, it sets up the threshold for crime scarcity based on a target crime rate (target_crime_rate), which adjusts the outcome distribution based on the desired crime rate within the dataset.
```{r}
# Noise set-up
noise_income <- 6000
noise_house_ownership <- 0.09
noise_crime <- 0.09

# Relationship of income to sensitive attributes and education
# linear relationship when set FALSE; else non-linearity introduced in form of interactions
non_linear_income <- FALSE

# missing data points set-up
# TRUE when dataset should contain missing values
missing_data <- FALSE

# Set-up for decision boundary
# Set to TRUE for non-linear decision boundary, FALSE for linear
non_linear <- TRUE 

# Scarcity of crime outcome
# Adjust threshold based on desired scarcity (e.g., target 10% crime rate)
target_crime_rate <- 0.1

# Flipping target label (crime) to introduce more noise
swap_rate <- 0.09
```

# Generate protected attributes
This section simulates protected attributes (gender, ethnicity, disability) for the dataset using predefined probabilities. These attributes will be used later to influence other variables in the dataset, such as education and income.
```{r}
# gender
gender <- as.factor(sample(c("Female", "Male"), n, replace = TRUE))

# ethnicity
ethnicity <- sample(c("Minority", "Majority"), n, replace = TRUE, prob = c(0.35, 0.65))

# disability
disability <- sample(c("Yes", "No"), n, replace = TRUE, prob = c(0.07, 0.93))

# Define the levels of education
education_levels <- c("No High School", "High School", "Some College", "Bachelor's", "Master's", "Doctorate")
```

# Education variable based on sensitive attributes
Here, the education levels are generated for each individual in the dataset based on their gender, ethnicity, and disability status. The probability of attaining different education levels varies depending on these protected attributes, simulating potential societal disparities.
```{r}
# Generate education levels with different probabilities for gender, ethnicity, and disability
education <- mapply(function(g, e, d) {
  if (d == "Yes") {
    sample(education_levels, size = 1, prob = c(0.4, 0.35, 0.15, 0.07, 0.02, 0.01))
  } else if (g == "Male" && e == "Majority") {
    sample(education_levels, size = 1, prob = c(0.264, 0.4, 0.2, 0.09, 0.04, 0.006))
  } else if (g == "Female" && e == "Majority") {
    sample(education_levels, size = 1, prob = c(0.21, 0.35, 0.25, 0.12, 0.06, 0.01))
  } else if (g == "Male" && e == "Minority") {
    sample(education_levels, size = 1, prob = c(0.35, 0.4, 0.15, 0.07, 0.02, 0.006))
  } else {
    sample(education_levels, size = 1, prob = c(0.3, 0.4, 0.2, 0.07, 0.02, 0.01))
  }
}, gender, ethnicity, disability)

education <- factor(education, levels = education_levels)
```

# Create a data frame with the simulated data
This section creates a data frame simulated_data containing the generated attributes and uses tables to summarize the distribution of education levels across gender, ethnicity, and disability groups.
```{r}
simulated_data <- data.frame(gender = gender, ethnicity = ethnicity, disability = disability, education = education)

table(simulated_data$gender, simulated_data$education)
table(simulated_data$ethnicity, simulated_data$education)
table(simulated_data$disability, simulated_data$education)
```
# Function for noise and probability handling
This code block defines a utility function, apply_noise_or_adjust_prob, which adds noise to a given variable, adjusts probabilities within specified bounds, and optionally applies a sigmoid transformation. It also allows for reproducibility by setting a random seed if provided.
```{r}
# Utility function to handle noise and probabilities
apply_noise_or_adjust_prob <- function(input_var, noise_sd = 0, prob_adjustment = 0, min_prob = 0, max_prob = 1, apply_sigmoid = FALSE, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  
  if (noise_sd > 0) {
    noise <- rnorm(length(input_var), mean = 0, sd = noise_sd)
    input_var <- input_var + noise
  }
  
  if (prob_adjustment != 0) {
    input_var <- input_var + prob_adjustment
    input_var <- pmin(pmax(input_var, min_prob), max_prob)
  }
  
  if (apply_sigmoid) {
    input_var <- 1 / (1 + exp(-input_var))
  }
  
  return(input_var)
}
```

# income variable based on sensitive attributes and education
Income is generated by combining a base income with various effects (gender, ethnicity, disability, education) and interaction effects. Noise is added to introduce variability, and summaries of income across different groups are provided to analyze disparities.
```{r}
# Base income setup
base_income <- 30000

# Gender effect (Females earn less than Males)
gender_effect <- ifelse(gender == "Male", 7000, -3000)

# Ethnicity effect (Minorities earn less, mitigated by high education)
ethnicity_effect <- ifelse(ethnicity == "Minority", -5000, 0)
education_mitigation <- ifelse(education %in% c("Master's", "Doctorate"), 5000, 0)
ethnicity_effect <- ethnicity_effect + education_mitigation
ethnicity_effect <- ifelse(ethnicity_effect == 5000, 0, ethnicity_effect)

# Disability effect (People with disabilities earn less)
disability_effect <- ifelse(disability == "Yes", -7000, 0)

# Education effect (Higher education corresponds to higher income)
education_effect <- sapply(education, function(e) {
  switch(as.character(e),
         "No High School" = 0,
         "High School" = 5000,
         "Some College" = 10000,
         "Bachelor's" = 20000,
         "Master's" = 35000,
         "Doctorate" = 50000)
})

if(non_linear_income == TRUE){
  # Interaction effects for non-linear relationships
  interaction_effect <- mapply(function(g, e, et, d) {
    interaction_val <- 0
    
    # Interaction between gender and education
    if (g == "Female") {
      interaction_val <- interaction_val - (as.numeric(e) * 500)
    }
    
    # Interaction between ethnicity and education
    if (et == "Minority") {
      interaction_val <- interaction_val - (as.numeric(e) * 300)
    }
    
    # Interaction between disability and education
    if (d == "Yes") {
      interaction_val <- interaction_val - (as.numeric(e) * 200)
    }
    
    # Non-linear effect (example: square of education effect)
    if (non_linear) {
      interaction_val <- interaction_val + (as.numeric(e)^2 * 100)
    }
    
    return(interaction_val)
  }, gender, education, ethnicity, disability)
}

# Calculate the income
income <- base_income + gender_effect + ethnicity_effect + disability_effect + education_effect

# Add non-linear relationship (interaction effect)
if(non_linear_income == TRUE){
  income <- income + interaction_effect
}

# Add noise to the income
income <- apply_noise_or_adjust_prob(income, noise_sd = noise_income, seed = 456)

# Add income to the data frame
simulated_data$income <- income

# Summaries
print("Income of whole population")
summary(simulated_data$income)
print("Income male population")
summary(simulated_data[simulated_data$gender == "Male",]$income)
print("Income female population")
summary(simulated_data[simulated_data$gender == "Female",]$income)
print("Income majority population")
summary(simulated_data[simulated_data$ethnicity == "Majority",]$income)
print("Income minority population")
summary(simulated_data[simulated_data$ethnicity == "Minority",]$income)
print("Income population with disability")
summary(simulated_data[simulated_data$disability == "Yes",]$income)
print("Income population without disability")
summary(simulated_data[simulated_data$disability == "No",]$income)
print("Income of non-disabled, majority man")
summary(simulated_data[simulated_data$gender == "Male" & simulated_data$disability == "No" & simulated_data$ethnicity == "Majority",]$income)
print("Income of non-disabled, majority woman")
summary(simulated_data[simulated_data$gender == "Female" & simulated_data$disability == "No" & simulated_data$ethnicity == "Majority",]$income)
print("Income of non-disabled, minority woman")
summary(simulated_data[simulated_data$gender == "Female" & simulated_data$disability == "No" & simulated_data$ethnicity == "Minority",]$income)

# Plot income distribution
hist(simulated_data$income)
```
# Former income variable generation
```{#r}
# Define base income, gender effect, education effect, ethnicity effect, disability effect, and interaction effects
base_income <- 30000

# gender effect
gender_effect <- ifelse(gender == "Male", 8500, 0)

# ethnicity effect
ethnicity_effect <- ifelse(ethnicity == "Minority", -5000, 0)

# disability effect
disability_effect <- ifelse(disability == "Yes", -7000, 0)

# education effect
education_effect <- sapply(education, function(e) {
  switch(as.character(e),
         "No High School" = 0,
         "High School" = 5000,
         "Some College" = 7500,
         "Bachelor's" = 10000,
         "Master's" = 20000,
         "Doctorate" = 30000)
})

# interaction effects (gender * education, ethnicity * education, gender * ethnicity * education, and disability interactions)
interaction_effect <- mapply(function(g, e, et, d) {
  if (d == "Yes" && g == "Female" && et == "Minority") {
    switch(as.character(e),
           "No High School" = -3000,
           "High School" = -2500,
           "Some College" = -2000,
           "Bachelor's" = -1500,
           "Master's" = -1000,
           "Doctorate" = -500)
  } else if (d == "Yes" && g == "Female") {
    switch(as.character(e),
           "No High School" = -2500,
           "High School" = -2000,
           "Some College" = -1500,
           "Bachelor's" = -1000,
           "Master's" = -500,
           "Doctorate" = 0)
  } else if (d == "Yes" && et == "Minority") {
    switch(as.character(e),
           "No High School" = -2500,
           "High School" = -2000,
           "Some College" = -1500,
           "Bachelor's" = -1000,
           "Master's" = -500,
           "Doctorate" = 0)
  } else if (g == "Female" && et == "Minority") {
    switch(as.character(e),
           "No High School" = -2000,
           "High School" = -1500,
           "Some College" = -1000,
           "Bachelor's" = -500,
           "Master's" = 0,
           "Doctorate" = 500)
  } else if (g == "Female") {
    switch(as.character(e),
           "No High School" = -1000,
           "High School" = -500,
           "Some College" = 0,
           "Bachelor's" = 500,
           "Master's" = 1000,
           "Doctorate" = 1500)
  } else if (et == "Minority") {
    switch(as.character(e),
           "No High School" = -1000,
           "High School" = -500,
           "Some College" = 0,
           "Bachelor's" = 500,
           "Master's" = 1000,
           "Doctorate" = 1500)
  } else {
    switch(as.character(e),
           "No High School" = 0,
           "High School" = 500,
           "Some College" = 1000,
           "Bachelor's" = 1500,
           "Master's" = 2000,
           "Doctorate" = 2500)
  }
}, gender, education, ethnicity, disability)


# Calculate the income
income <- base_income + gender_effect + ethnicity_effect + disability_effect + education_effect + interaction_effect

# Add noise to the income using utility function
income <- apply_noise_or_adjust_prob(income, noise_sd = noise_income, seed = 456)

# Add income to the data frame
simulated_data$income <- income

# Summaries
summary(simulated_data$income)
summary(simulated_data[simulated_data$gender == "Male",])
summary(simulated_data[simulated_data$gender == "Female",])
summary(simulated_data[simulated_data$ethnicity == "Majority",])
summary(simulated_data[simulated_data$ethnicity == "Minority",])
summary(simulated_data[simulated_data$disability == "Yes",])
summary(simulated_data[simulated_data$disability == "No",])

hist(simulated_data$income)
```

# Test relationships to income
A generalized linear model (GLM) is fitted to assess the relationship between income and the other variables in the dataset.
```{r}
model_income <- glm(income ~ ., data = simulated_data)
summary(model_income)
```

# House ownership variable
This section generates a binary house ownership variable based on education level and adjusted by income. Probabilities are defined, noise is added, and the results are summarized in tables to explore relationships between house ownership and other variables.
```{r}
# Define base probabilities for house ownership by education
base_probabilities <- sapply(education, function(e) {
  switch(as.character(e),
         "No High School" = 0.1,
         "High School" = 0.2,
         "Some College" = 0.3,
         "Bachelor's" = 0.5,
         "Master's" = 0.6,
         "Doctorate" = 0.7)
})

# Adjust probabilities based on income
income_effect <- (income - mean(income)) / sd(income) * 0.1
ownership_probabilities <- base_probabilities + income_effect

# Adjust probabilities for house ownership using utility function
ownership_probabilities <- apply_noise_or_adjust_prob(ownership_probabilities, noise_sd = noise_house_ownership, min_prob = 0, max_prob = 1, seed = 789)

# Determine house ownership using a threshold
set.seed(101112)  # Setting seed for reproducibility
house_ownership <- ifelse(runif(n) < ownership_probabilities, 1, 0)

# Add house ownership to the data frame
simulated_data$house_ownership <- house_ownership

# inspect house ownership
table(simulated_data$gender, simulated_data$house_ownership)
table(simulated_data$house_ownership, simulated_data$education)
table(simulated_data$disability, simulated_data$house_ownership)
table(simulated_data$ethnicity, simulated_data$house_ownership)
```
# Function to generate new variable
A custom function generate_new_variable is defined to create new variables based on existing ones. This function supports various functional forms (linear, exponential, power, etc.) and can produce binary outcomes based on specified cut-off criteria.
```{r}
generate_new_variable <- function(data, existing_vars, coefficients, noise_sd, functional_form, binary = FALSE, cut_off = NULL, non_linear = FALSE) {
  
  # Ensure existing_vars and coefficients are of the same length
  if(length(existing_vars) != length(coefficients)) {
    stop("The number of existing variables and coefficients must be the same.")
  }
  
  # Extract the relevant columns from the data and convert to numeric
  vars <- data[, existing_vars]
  vars <- as.data.frame(lapply(vars, as.numeric))
  
  # Calculate the linear combination of variables and coefficients
  linear_combination <- as.matrix(vars) %*% coefficients
  
  # Add noise
  set.seed(123)  # Ensure reproducibility
  noise <- rnorm(nrow(data), mean = 0, sd = noise_sd)
  
  # Apply non-linear transformations if non_linear is TRUE
  if (non_linear) {
    if (functional_form == "linear") {
      new_var <- linear_combination^2 + noise  # Example: squaring the linear combination
    } else if (functional_form == "exp_linear") {
      new_var <- exp(linear_combination^2 + noise)
    } else if (functional_form == "power") {
      new_var <- (rowSums(vars^coefficients))^2 + noise
    } else if (functional_form == "exp_power") {
      new_var <- exp((rowSums(vars^coefficients))^2 + noise)
    } else if (functional_form == "sigmoid") {
      new_var <- 1 / (1 + exp(-linear_combination^2 - noise))
    } else if (functional_form == "log") {
      new_var <- rowSums(coefficients * log(abs(vars)^2 + 1)) + noise
    } else {
      stop("Invalid functional form selected.")
    }
  } else {
    # Keep the standard functional forms if non_linear is FALSE
    if (functional_form == "linear") {
      new_var <- linear_combination + noise
    } else if (functional_form == "exp_linear") {
      new_var <- exp(linear_combination + noise)
    } else if (functional_form == "power") {
      new_var <- rowSums(vars^coefficients) + noise
    } else if (functional_form == "exp_power") {
      new_var <- exp(rowSums(vars^coefficients) + noise)
    } else if (functional_form == "sigmoid") {
      new_var <- 1 / (1 + exp(-(linear_combination + noise)))
    } else if (functional_form == "log") {
      new_var <- rowSums(coefficients * log(abs(vars) + 1)) + noise  # +1 to avoid log(0)
    } else {
      stop("Invalid functional form selected.")
    }
  }
  
  # Convert to binary if specified
  if (binary) {
    if (is.null(cut_off)) {
      stop("A cut-off method (mean or median) must be specified for binary conversion.")
    } else if (cut_off == "mean") {
      new_var <- ifelse(new_var > mean(new_var), 1, 0)
    } else if (cut_off == "median") {
      new_var <- ifelse(new_var > median(new_var), 1, 0)
    } else {
      stop("Invalid cut-off method. Choose 'mean' or 'median'.")
    }
  }
  
  return(new_var)
}


# Example:
# new_var <- generate_new_variable(data = simulated_data, 
#                                  existing_vars = c("income", "education"),
#                                  coefficients = c(0.5, 1.5), 
#                                  noise_sd = 10, 
#                                  functional_form = "linear", 
#                                  binary = TRUE, 
#                                  cut_off = "mean",
#                                  non_linear = non_linear)

```

# Initialise new variables
New variables are generated using the custom function defined earlier. Each new variable is added to the dataset, and its relationship with existing variables is analyzed using GLMs.
```{r}
# 1st new variable
new_var <- generate_new_variable(data = simulated_data, 
                                  existing_vars = c("income", "education"),
                                  coefficients = c(0.5, 1.5), 
                                  noise_sd = 10, 
                                  functional_form = "linear", 
                                  binary = TRUE, 
                                  cut_off = "mean",
                                  non_linear = non_linear)

simulated_data$new_var <- new_var

var_glm <- glm(new_var ~ ., data = simulated_data)
summary(var_glm)

# 2nd new variable
new_var2 <- generate_new_variable(data = simulated_data, 
                                  existing_vars = c("income", "education", "new_var"),
                                  coefficients = c(3.6, 0.7, 2.3), 
                                  noise_sd = 3, 
                                  functional_form = "log", 
                                  binary = FALSE, 
                                  cut_off = NULL,
                                  non_linear = non_linear)

simulated_data$new_var2 <- new_var2

var_glm <- glm(new_var2 ~ ., data = simulated_data)
summary(var_glm)

# 3rd variable
new_var3 <- generate_new_variable(data = simulated_data, 
                                  existing_vars = c("house_ownership", "new_var", "new_var2"),
                                  coefficients = c(0.12, 0.7, 0.03), 
                                  noise_sd = 4, 
                                  functional_form = "sigmoid", 
                                  binary = TRUE, 
                                  cut_off = "median",
                                  non_linear = non_linear)

simulated_data$new_var3 <- new_var3

var_glm <- glm(new_var3 ~ ., data = simulated_data)
summary(var_glm)

# 4th variable
new_var4 <- generate_new_variable(data = simulated_data, 
                                  existing_vars = c("house_ownership","income", "new_var3"),
                                  coefficients = c(-6.2, -0.0004, -13), 
                                  noise_sd = 6, 
                                  functional_form = "linear", 
                                  binary = FALSE, 
                                  cut_off = NULL,
                                  non_linear = non_linear)

simulated_data$new_var4 <- new_var4

var_glm <- glm(new_var4 ~ ., data = simulated_data)
summary(var_glm)

head(simulated_data)
```

# Generate crime variable
A crime variable is simulated based on a combination of the previously generated variables. A logistic function (sigmoid) is applied to calculate the probability of crime, and a binary crime variable is generated based on a specified threshold.
```{r}
# Sigmoid function
sigmoid <- function(x){
  1 / (1 + exp(-x))
}

# Linear Combination
linear_combination <- -0.2 * as.numeric(simulated_data$education) +
                      -0.00003 * simulated_data$income +
                      -1.7 * simulated_data$house_ownership +
                      -0.7 * simulated_data$new_var +
                      0.0007 * simulated_data$new_var2 +
                      0.065 * simulated_data$new_var3 +
                      0.0009 * simulated_data$new_var4

# Adjustments for non-linear decision boundary
if(non_linear == TRUE){
  # Optional Non-Linear Terms
  non_linear_component <- -0.05 * (simulated_data$income)^2 + 
                          0.02 * (simulated_data$new_var * simulated_data$new_var2) + 
                          -0.03 * (simulated_data$income * as.numeric(simulated_data$education))
  
  # Combine linear and non-linear components
  combined_effect <- linear_combination + non_linear_component
} else {
  combined_effect <- linear_combination
}

# Normalize the combined effect and convert back to vector
combined_effect <- as.vector(scale(combined_effect, center = TRUE, scale = TRUE)) # Ensure it's a vector

# Calculate crime probability using sigmoid function
crime_prob <- sigmoid(combined_effect)

# Introduce noise
noise <- rnorm(length(crime_prob), mean = 0, sd = noise_crime)
crime_prob <- pmin(pmax(crime_prob + noise, 0), 1)

# Dynamic Threshold Adjustment
# Adjust threshold based on desired scarcity in set-up at beginning of the R markdown (e.g., target 10% crime rate)
scarcity_threshold <- quantile(crime_prob, 1 - target_crime_rate, na.rm = TRUE)

# Generate binary crime variable based on adjusted threshold
simulated_data$crime <- ifelse(crime_prob > scarcity_threshold, 1, 0)

# Summarize the new crime variable
summary(simulated_data$crime)
table(simulated_data$crime)
```
# Randomly distort crime classification
This code block randomly distorts a specified proportion of the crime classification outcomes in the dataset, effectively flipping some outcomes from 1 to 0 or from 0 to 1. This can be used to simulate noise, errors, or bias in the classification process for experimental or testing purposes.
```{r}
# select indices
swap_indices <- sample(1:n, size = round(n * swap_rate, 0), replace = FALSE)

# re-classify cases to opposite outcome
simulated_data$crime[swap_indices] <- ifelse(simulated_data$crime[swap_indices] == 1, 0, 1)
```

# Creating missing values
This section introduces missing values into the dataset using different mechanisms:

- Completely Random Missing Values: Introduces missing values at random.
- Distribution-Based Missing Values: Missingness is based on the distribution of a specific variable.
- Dependent Missing Values: Missingness is influenced by other variables in the dataset.
- Sensitive-Dependent Missing Values: Missingness is influenced by sensitive attributes like gender and ethnicity.

These scenarios help in simulating real-world data conditions where missingness is not purely random.
```{r}
create_missing_values <- function(data, var_names, proportion = NULL, num_missing = NULL, 
                                  type = c("random", "distribution", "dependent", "sensitive"),
                                  dist_params = list(), 
                                  open_var = NULL, 
                                  sensitive_var = NULL) {
  
  # Convert proportion to num_missing if proportion is given
  if (!is.null(proportion)) {
    num_missing <- floor(proportion * nrow(data))
  }
  
  if (is.null(num_missing)) {
    stop("Either proportion or num_missing must be provided.")
  }
  
  # Case 1: Completely Random Missing Values
  if (type == "random") {
    for (var_name in var_names) {
      missing_indices <- sample(1:nrow(data), num_missing, replace = FALSE)
      data[missing_indices, var_name] <- NA
    }
  }
  
  # Case 2: Missing Values Based on a Distribution
  if (type == "distribution") {
    if (length(var_names) != 1) {
      stop("For distribution-based missingness, provide exactly one variable.")
    }
    
    var_name <- var_names[1]
    if (is.numeric(data[[var_name]])) {
      # Quantile-based probability for continuous variables
      quantiles <- quantile(data[[var_name]], probs = c(0.25, 0.5, 0.75, 1.0), na.rm = TRUE)
      probabilities <- c(0.1, 0.3, 0.5, 0.7)  # Example: Increase missingness probability across quantiles
      data$missing_prob <- cut(data[[var_name]], breaks = c(-Inf, quantiles), labels = probabilities, include.lowest = TRUE)
      data$missing_prob <- as.numeric(as.character(data$missing_prob))
    } else {
      # Category-based probability for categorical variables
      categories <- levels(factor(data[[var_name]]))
      probabilities <- seq(0.1, 0.7, length.out = length(categories))
      prob_map <- setNames(probabilities, categories)
      data$missing_prob <- prob_map[as.character(data[[var_name]])]
    }
    data$missing_prob[is.na(data$missing_prob)] <- 0
    missing_indices <- order(-data$missing_prob)[1:num_missing]
    data[missing_indices, var_name] <- NA
    data$missing_prob <- NULL  # Clean up
  }
  
  # Case 3: Missing Values Dependent on Other Variables
  if (type == "dependent") {
    if (is.null(open_var) || length(var_names) != 1) {
      stop("For dependent missingness, provide exactly one target variable and at least one open_var.")
    }
    
    var_name <- var_names[1]
    linear_combination <- 0
    for (var in open_var) {
      linear_combination <- linear_combination + as.numeric(data[[var]])
    }
    probabilities <- 1 / (1 + exp(-linear_combination))  # Sigmoid function to calculate probabilities
    data$missing_prob <- probabilities
    missing_indices <- order(-data$missing_prob)[1:num_missing]
    data[missing_indices, var_name] <- NA
    data$missing_prob <- NULL  # Clean up
  }
  
  # Case 4: Missing Values Dependent on Sensitive Characteristics
  if (type == "sensitive") {
    if (is.null(sensitive_var) || length(var_names) != 1) {
      stop("For sensitive missingness, provide exactly one target variable and at least one sensitive_var.")
    }
    
    var_name <- var_names[1]
    linear_combination <- 0
    for (var in sensitive_var) {
      linear_combination <- linear_combination + as.numeric(data[[var]])
    }
    probabilities <- 1 / (1 + exp(-linear_combination))  # Sigmoid function to calculate probabilities
    data$missing_prob <- probabilities
    missing_indices <- order(-data$missing_prob)[1:num_missing]
    data[missing_indices, var_name] <- NA
    data$missing_prob <- NULL  # Clean up
  }
  
  return(data)
}


if(missing_data == TRUE){
  # Adding random missing values
  simulated_data <- create_missing_values(simulated_data, var_names = c("income", "education"), 
                                        proportion = 0.02, type = "random")

  # Adding distribution-based missing values
  simulated_data <- create_missing_values(simulated_data, var_names = "income", 
                                          num_missing = 100, type = "distribution")
  
  # Adding dependent missing values
  simulated_data <- create_missing_values(simulated_data, var_names = "new_var4", 
                                          num_missing = 75, type = "dependent", 
                                          open_var = c("income", "education"))
  
  # Adding sensitive-dependent missing values
  simulated_data <- create_missing_values(simulated_data, var_names = "house_ownership", 
                                          num_missing = 75, type = "sensitive", 
                                          sensitive_var = c("gender", "ethnicity"))
  
  # Overview of NA values
  table(is.na(simulated_data$income))
}

```

