---
title: "Threshold Plots"
author: "Leona"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*Important:* Run analysis of simulation first, to ensure probabilities for the model are stored in global environment 

# Load library
```{r}
library(tidyverse)
library(gridExtra)
library(cowplot)
```

# Initial set-up
This chunk initializes the key variables:
*current_model:* Sets the model's name for labeling.
*probabilities:* Represents the predicted probabilities from the test data.
*data_plots:* Contains the data used for plotting, including the outcomes (e.g., crime).
*Vulnerable and privileged groups* are defined, such as gender and ethnicity categories. These groupings are used later to analyze fairness metrics.
```{r}
# Set current model for labelling of the plots
current_model <- "XGBoost"

# Predicted probabilities
probabilities <- probabilities_test

# used data
data_plots <- data_test

# Define vulnerable group (contrasted group 1)
gender_vulnerable <- "Female"
disability_vulnerable <- NULL
ethnicity_vulnerable <- "Minority"

# Define privileged group (contrasted group 2)
gender_privileged <- "Male"
disability_privileged <- NULL
ethnicity_privileged <- "Majority"

# Best model
best_model <- FALSE
```

# Set-up to store values necessary for later plotting
This section sets up containers to store values for three fairness measures (False Positive Rate (FPR), False Negative Rate (FNR), and the proportion of favorable outcomes) across different thresholds (from 0.05 to 0.95).
```{r}
# Threshold sequence
threshold <- seq(0.05, 0.95, 0.05)

# Empty containers for fairness metrics
# - 1. False positive rate
bench_FPR <- c()
vulnerable_FPR <- c()
privileged_FPR <- c()

# - 2. false negative rate
bench_FNR <- c()
vulnerable_FNR <- c()
privileged_FNR <- c()

# - 3. proportion of favourable outcome (0 - no crime) for demographic parity
bench_proportion <- c()
vulnerable_proportion <- c()
privileged_proportion <- c()
```

# Function to calculate the three fairness measures for the defined groups 
This function, calculate_metrics_for_group(), calculates three key metrics:

*False Positive Rate (FPR):* The probability of predicting a crime when no crime occurred.
*False Negative Rate (FNR):* The probability of predicting no crime when a crime did occur.
*Proportion of Favorable Outcomes:* The percentage of predictions where no crime occurred.
It calculates these metrics across multiple thresholds and groups (vulnerable and privileged) based on gender, ethnicity, or disability.
```{r}
# Function to calculate fairness metrics for a group
calculate_metrics_for_group <- function(data, probabilities, threshold, gender = NULL, disability = NULL, ethnicity = NULL) {
  
  # Filter data based on group characteristics
  if (!is.null(gender)) {
    data <- data[data$gender == gender,]
  }
  
  if (!is.null(disability)) {
    data <- data[data$disability == disability,]
  }
  
  if (!is.null(ethnicity)) {
    data <- data[data$ethnicity == ethnicity,]
  }
  
  # Filter the probabilities to match the filtered data
  probabilities <- probabilities[rownames(data_plots) %in% rownames(data)]
  
  FPR <- c()
  FNR <- c()
  proportion_favorable <- c()
  
  # Calculate metrics for each threshold
  for (t in threshold) {
    predictions <- ifelse(probabilities >= t, 1, 0)
    
    # Ensure that both 'data$crime' and 'predictions' have the same length
    if (length(data$crime) == length(predictions)) {
      cm <- table(data$crime, predictions)
      
      TN <- ifelse("0" %in% rownames(cm) && "0" %in% colnames(cm), cm["0", "0"], 0)
      FP <- ifelse("0" %in% rownames(cm) && "1" %in% colnames(cm), cm["0", "1"], 0)
      FN <- ifelse("1" %in% rownames(cm) && "0" %in% colnames(cm), cm["1", "0"], 0)
      TP <- ifelse("1" %in% rownames(cm) && "1" %in% colnames(cm), cm["1", "1"], 0)
      
      # FPR = FP / (FP + TN)
      if (TN + FP > 0) {
        FPR <- c(FPR, FP / (FP + TN))
      } else {
        FPR <- c(FPR, NA)
      }
      
      # FNR = FN / (FN + TP)
      if (FN + TP > 0) {
        FNR <- c(FNR, FN / (FN + TP))
      } else {
        FNR <- c(FNR, NA)
      }
      
      # Proportion of favorable outcome (0 = no crime)
      proportion_favorable <- c(proportion_favorable, sum(predictions == 0) / length(predictions))
      
    } else {
      stop("The lengths of the filtered data and predictions do not match.")
    }
  }
  
  return(list(FPR = FPR, FNR = FNR, proportion_favorable = proportion_favorable))
}
```

# Calculate metrics for defined groups and benchmark
This chunk computes the fairness metrics for three groups:

Benchmark: Overall metrics for the entire population.
Vulnerable Group: Metrics specific to the vulnerable group (e.g., Female + Minority).
Privileged Group: Metrics specific to the privileged group (e.g., Male + Majority).
These computed values are stored in vectors (e.g., bench_FPR, vulnerable_FPR, etc.) for later plotting.
```{r}
# Calculate metrics for benchmark (all data)
benchmark_metrics <- calculate_metrics_for_group(data_plots, probabilities, threshold)

# Calculate metrics for vulnerable group
vulnerable_metrics <- calculate_metrics_for_group(data_plots, probabilities, threshold, gender_vulnerable, disability_vulnerable, ethnicity_vulnerable)

# Calculate metrics for privileged group
privileged_metrics <- calculate_metrics_for_group(data_plots, probabilities, threshold, gender_privileged, disability_privileged, ethnicity_privileged)

# Store the metrics for plotting
bench_FPR <- benchmark_metrics$FPR
bench_FNR <- benchmark_metrics$FNR
bench_proportion <- benchmark_metrics$proportion_favorable

vulnerable_FPR <- vulnerable_metrics$FPR
vulnerable_FNR <- vulnerable_metrics$FNR
vulnerable_proportion <- vulnerable_metrics$proportion_favorable

privileged_FPR <- privileged_metrics$FPR
privileged_FNR <- privileged_metrics$FNR
privileged_proportion <- privileged_metrics$proportion_favorable
```

# Plots for ratio to benchmark of FPR and FNR, next to each other
This plot shows the ratio of the FPR and FNR of both the vulnerable and privileged groups to the overall benchmark, across different thresholds.

Interpretation: A line above 1 indicates the group has a higher rate (FPR or FNR) compared to the benchmark, meaning the model is more likely to make errors for that group. If both groups hover around 1, the model treats both groups similarly in terms of false positives or false negatives.
```{r}
# Calculate the ratios
vulnerable_FPR_ratio <- vulnerable_FPR / bench_FPR
privileged_FPR_ratio <- privileged_FPR / bench_FPR

vulnerable_FNR_ratio <- vulnerable_FNR / bench_FNR
privileged_FNR_ratio <- privileged_FNR / bench_FNR

# Create a data frame for plotting
plot_data <- data.frame(
  Threshold = rep(threshold, times = 4),
  Metric = rep(c("FPR Ratio", "FNR Ratio"), each = length(threshold) * 2),
  Group = rep(c(paste("Vulnerable Group:", gender_vulnerable, ethnicity_vulnerable, disability_vulnerable),
                paste("Privileged Group:", gender_privileged, ethnicity_privileged, disability_privileged)), 
              each = length(threshold)),
  Value = c(vulnerable_FPR_ratio, privileged_FPR_ratio, 
            vulnerable_FNR_ratio, privileged_FNR_ratio)
)

# Best model: store for later usage
if(best_model == TRUE) {
  best_plot_data <- plot_data
}

# Plot
ggplot(plot_data, aes(x = Threshold, y = Value, color = Group)) +
  geom_line(size = 1) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey") +
  facet_wrap(~ Metric, scales = "free_y") +
  labs(title = paste(current_model, "FPR and FNR Ratios to Benchmark"),
       x = "Threshold",
       y = "Ratio to Benchmark",
       color = "Group") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "bottom")
```
# Ratio to benchmark of FPR and FNR in one plot
This plot is similar to the previous one but combines the FPR and FNR ratios into a single graph with different line types for each metric.

Interpretation: This allows for a clearer comparison of the relative differences in FPR and FNR across the vulnerable and privileged groups. Again, lines above 1 signify worse performance compared to the benchmark, highlighting potential bias.
```{r}
# Plot
ggplot(plot_data, aes(x = Threshold, y = Value, color = Group, linetype = Metric)) +
  geom_line(size = 1) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey") +
  labs(title = paste(current_model, "FPR and FNR Ratios to Benchmark"),
       x = "Threshold",
       y = "Ratio to Benchmark",
       color = "Group",
       linetype = "Metric") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "right")
```
# FNR and FPR for both groups and benchmark
Here, FPR and FNR for both vulnerable and privileged groups, alongside the benchmark, are plotted against the threshold.

Interpretation: This plot allows us to see the absolute values of the false positive and false negative rates for each group, rather than relative to the benchmark. A significant difference between groups would indicate potential unfairness or bias, with one group experiencing more false positives or false negatives at varying thresholds.
```{r}
# Combine FPR and FNR data into a single data frame
combined_data <- data.frame(
  Threshold = rep(threshold, 6),
  Metric = rep(c("FPR", "FNR"), each = length(threshold) * 3),
  Group = rep(c(paste("Benchmark"),
                paste("Vulnerable Group:", gender_vulnerable, ethnicity_vulnerable, disability_vulnerable),
                paste("Privileged Group:", gender_privileged, ethnicity_privileged, disability_privileged)), 
              times = 2, each = length(threshold)),
  Value = c(bench_FPR, vulnerable_FPR, privileged_FPR, bench_FNR, vulnerable_FNR, privileged_FNR)
)

# Best model: store for later usage
if(best_model == TRUE) {
  best_combined_data <- combined_data
}

# Plot the combined data
combined_plot <- ggplot(combined_data, aes(x = Threshold, y = Value, color = Group, linetype = Metric)) +
  geom_line(size = 1) +
  labs(title = paste(current_model, "- FPR and FNR Combined"),
       x = "Threshold",
       y = "Rate") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "right")

# Display the final plot
print(combined_plot)

```

# Demographic parity plot (proportion of favourable (crime == 0) outcome)
This plot shows the proportion of favorable outcomes (i.e., predictions of no crime) for the benchmark, vulnerable, and privileged groups.

Interpretation: This plot examines demographic parity: whether each group receives favorable outcomes at similar rates. If there are large differences, this suggests that one group is being treated more favorably than the other, raising fairness concerns.
```{r}
# Combine the proportion data into a single data frame
proportion_data <- data.frame(
  Threshold = threshold,
  Group = rep(c(paste("Benchmark"),
                paste("Vulnerable Group:", gender_vulnerable, ethnicity_vulnerable, disability_vulnerable),
                paste("Privileged Group:", gender_privileged, ethnicity_privileged, disability_privileged)), 
              each = length(threshold)),
  Proportion = c(bench_proportion, vulnerable_proportion, privileged_proportion)
)

if(best_model == TRUE) {
  best_proportion_data <- proportion_data
}

# Plot the proportion of favorable outcomes
proportion_plot <- ggplot(proportion_data, aes(x = Threshold, y = Proportion, color = Group)) +
  geom_line(size = 1) +
  labs(title = paste(current_model, "- Proportion of Favorable Outcomes"),
       x = "Threshold",
       y = "Proportion of Favorable Outcomes (No Crime)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "right")

# Display the final plot
print(proportion_plot)
```

